
\documentclass[review,12pt]{elsarticle}

\usepackage{amsmath, amssymb, amsthm, bm}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{graphicx}
\graphicspath{{../figures/}}

\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{remark}{Remark}

\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathbb{V}\mathrm{ar}}
\newcommand{\Prob}{\mathbb{P}}

\journal{Journal of the American Statistical Association}

\begin{document}

\begin{frontmatter}

\title{Closed-form Variational Inference for Count Statistical Agent-Based Models}

\author{Xulin Pan\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{xxp5066@psu.edu}
\address{Pennsylvania State University,
         University Park, PA 16802, USA}

\begin{abstract}
Bayesian inference for count-valued agent-based models (ABMs) is impeded by
the intractability of the transition likelihood: when latent abundance evolves
through a Binomial survival plus Poisson immigration convolution, neither the
log-PMF nor the exact evidence lower bound (ELBO) has a closed form.
We show that the Poisson moment generating function (MGF) resolves this
obstacle by delivering a \emph{fully closed-form} surrogate ELBO.  Under a
Poisson mean-field variational family, every expectation---survival
probabilities, immigration rates, and binary detection probabilities---is
computed analytically via the MGF with no quadrature or Monte Carlo, defining
a surrogate $\tilde{\mathcal{L}}$ with three explicitly bounded approximation
sources (Theorem~\ref{thm:elbo}).  Each coordinate-ascent update of the
resulting \emph{structured mean-field variational Bayes} (SMFVB) algorithm is
likewise closed-form, and the iterates are monotonically non-decreasing and
converge geometrically at rate $\rho\le1/2$, inherited from the Foster--Lyapunov drift condition of the
latent chain (Theorem~\ref{thm:vi-converge}).  We further establish that all
three approximation errors vanish as mean latent abundance grows, so the
SMFVB posterior is consistent in the large-population regime
(Corollary~\ref{cor:elbo-gap}), and that the dual-channel observation
model---fusing NegBin count surveys with Bernoulli occupancy detections---is
locally identifiable under mild temporal abundance variation
(Proposition~\ref{prop:identif-dual}).  In a simulation study (100
replicates, $m=100$, $T=15$), SMFVB achieves 95\% credible-interval coverage
within 4 percentage points of particle MCMC with a $22\times$ wall-time
reduction (4~min vs.\ 92~min).
\end{abstract}

\begin{keyword}
closed-form variational inference \sep
moment generating function \sep
evidence lower bound \sep
coordinate-ascent variational inference \sep
count statistical agent-based model \sep
negative binomial \sep
spatio-temporal model \sep
identifiability
\end{keyword}

\end{frontmatter}

\section{Introduction}\label{sec:intro}

Agent-based models (ABMs) express population dynamics through explicit
mechanistic rules at the level of individual agents or local groups.  When the
quantities of interest are non-negative integer counts---species abundances,
disease case loads, cell populations---the natural latent state is a
count-valued field $N_{i,t}\in\mathbb{Z}_{\ge0}$ evolving through stochastic
rules for local survival, neighbourhood dispersal, and long-distance arrival
\citep{hooten2010,heard2014}.  The resulting class of \emph{count statistical
ABMs} (count ABMs) accommodates open population dynamics with
density regulation, spatial connectivity, and multi-source observation, extending
the N-mixture family of \citet{dail2011} and the integrated data model
framework of \citet{verhoef2025} to mechanistic, agent-based transition kernels.

\paragraph{The inference challenge.}
Exact Bayesian inference for count ABMs is intractable.  The latent count at
each cell and time is the sum of three conditionally independent components:
\[
  N_{i,t} = S_{i,t} + I_{i,t} + L_{i,t},
\]
where $S_{i,t}\sim\mathrm{Binomial}(N_{i,t-1},p_i)$ captures density-regulated
survival, $I_{i,t}\sim\mathrm{Poisson}(\Lambda_{i,t})$ captures neighbourhood
immigration, and $L_{i,t}\sim\mathrm{Poisson}(\psi_i)$ captures long-distance
arrivals.  The Binomial--Poisson convolution has no closed-form probability
mass function, so the process log-likelihood
$\log p(N_{i,t}\mid\mathbf{N}_{t-1})$ cannot be computed exactly.  This makes
the evidence lower bound (ELBO)---the standard variational objective---analytically
inaccessible, and renders gradient-based variational methods inapplicable without
numerical approximation.

Existing approaches circumvent the intractability at the cost of either
computational efficiency or inferential exactness.  Particle MCMC
\citep{andrieu2010} targets the exact posterior but requires $O(T\cdot P)$
model evaluations per MCMC step, where $P$ is the particle count, making it
prohibitive for moderate $m$ and $T$.  Approximate Bayesian computation (ABC,
\citealt{beaumont2009}) avoids likelihood evaluation entirely but requires careful
summary statistic choice and yields only an approximate posterior.  Neither
provides a closed-form variational objective that can be optimised by
coordinate-ascent iteration.

\paragraph{Our approach: closed-form ELBO via the Poisson MGF.}
We show that replacing the intractable Binomial--Poisson process density with a
Poisson surrogate---justified by the Le Cam total-variation bound
$d_{\mathrm{TV}}(\mathrm{Binom}(n,p),\,\mathrm{Pois}(np))\le p$---converts
every expectation in the ELBO into a closed-form expression via the Poisson
moment generating function (MGF).  Survival expectations, binary detection
probabilities, and neighbourhood immigration rates are all obtained
analytically; no numerical integration, Monte Carlo sampling, or black-box
gradient estimator is required at any stage of the optimisation.

The closed-form ELBO enables a \emph{structured mean-field variational Bayes}
(SMFVB) algorithm: a coordinate-ascent (CAVI) scheme over a Poisson
mean-field family in which each iterate $\mu_{i,t}^\star$ is updated in
closed form.  The first-order optimality condition reveals this update as a
\emph{multiplicatively corrected conditional-mean forward pass}---the Bayesian
analogue of the Kalman prediction--correction step for this nonlinear,
non-Gaussian, discrete-state model---and the geometric convergence rate
$\rho\le1/2$ follows directly from the same closed-form structure.

\paragraph{Contributions.}
This paper makes four contributions:
\begin{enumerate}[label=(\roman*)]
\item \textbf{Closed-form surrogate ELBO} (Theorem~\ref{thm:elbo},
  \S\ref{sec:elbo-terms}): the Poisson MGF yields exact closed-form
  expressions for every expectation in the ELBO under a Poisson mean-field
  family, with three explicitly bounded approximation sources and no
  numerical integration.
\item \textbf{Closed-form CAVI algorithm} (\S\ref{sec:smfvb-alg}): each
  coordinate-ascent update is analytic; the iterates converge monotonically
  and geometrically at rate $\rho\le1/2$
  (Theorem~\ref{thm:vi-converge}, \S\ref{sec:vi-convergence}).
\item \textbf{Consistency and identifiability} (Corollary~\ref{cor:elbo-gap},
  Proposition~\ref{prop:identif-dual}, \S\ref{sec:identif-val}): the surrogate
  ELBO error vanishes as mean abundance grows; the dual-channel observation
  model is locally identifiable given $T\ge2$ with temporal abundance variation.
\item \textbf{Simulation evidence} (\S\ref{sec:simulation}): 100 replicates
  at $m=100$, $T=15$ show 95\% credible-interval coverage within 4 pp of
  particle MCMC at $22\times$ lower wall time.
\end{enumerate}

\paragraph{Organisation.}
Section~\ref{sec:process} specifies the count ABM.
Section~\ref{sec:vi} develops the MGF surrogate ELBO and the SMFVB algorithm.
Section~\ref{sec:identif-val} establishes convergence, consistency, and
identifiability.
Section~\ref{sec:simulation} presents the simulation study.
Section~\ref{sec:discussion} concludes.

\section{Notation}

Let \(S=\{1,\ldots,m\}\) denote spatial cells and \(t=1,\ldots,T\) denote discrete time.
The latent abundance at cell \(i\) and time \(t\) is

\[
N_{i,t}\in\mathbb{Z}_{\ge 0}.
\]

Let \(\mathcal{N}_i\) denote the neighborhood of cell \(i\), such as a queen or rook
neighborhood on a spatial lattice. Let \(x_i\in\mathbb{R}^p\) denote environmental
covariates and \(e_{i,t}\in\mathbb{R}^q\) denote survey-effort covariates.

\begin{table}[h!]
\centering
\caption{Core notation.}
\begin{tabular}{ll}
\toprule
Symbol & Meaning \\
\midrule
\(N_{i,t}\) & latent count in cell \(i\) at time \(t\) \\
\(Y^{(C)}_{i,t}\) & count-survey observation (NegBin channel) \\
\(Y^{(P)}_{i,t}\) & binary occupancy observation (Bernoulli channel) \\
\(S_{i,t}\) & surviving individuals \\
\(I_{i,t}\) & neighbourhood immigrants \\
\(L_{i,t}\) & long-distance arrivals \\
\(\phi_i\) & baseline persistence probability \\
\(K_i\) & local carrying capacity \\
\(\psi_i\) & long-distance arrival rate \\
\(\rho_{i,t}\) & count-survey detection probability \\
\(r\) & NegBin overdispersion parameter \\
\(\kappa_{i,t}\) & occupancy detection intensity \\
\(\eta_{i,t}\) & occupancy false-positive probability \\
\(\lambda\) & variational parameters $\{\mu_{i,t},\mathbf{m}_\alpha,S_\alpha,\{\nu_k\}\}$ \\
\(\mu_{i,t}\) & variational mean for $N_{i,t}$ under $q$ \\
\(\tilde{\mathcal{L}}\) & surrogate ELBO \\
\(\mathcal{Q}\) & Poisson mean-field variational family \\
\bottomrule
\end{tabular}
\end{table}

\section{Latent Process Model}\label{sec:process}

The latent count evolves through three additive components:

\[
N_{i,t}=S_{i,t}+I_{i,t}+L_{i,t},\qquad t=2,\ldots,T.
\]

Conditional on the previous latent field \(\bm N_{t-1}=(N_{1,t-1},\ldots,N_{m,t-1})\),
the three components are assumed conditionally independent.

\subsection{Density-Regulated Local Persistence}

Surviving individuals are modeled as

\[
S_{i,t}\mid N_{i,t-1},\phi_i,K_i
\sim
\mathrm{Binomial}\left(N_{i,t-1},p^S_{i,t-1}\right),
\]

where

\[
p^S_{i,t-1}
=
\phi_i\exp\left(-\frac{N_{i,t-1}}{K_i}\right).
\]

This is a Ricker-type density-regulated persistence model \citep{turchin1998}.
Empirical evidence that density-dependent dispersal and survival improve near-term
range-shift forecasts over climate-only models is provided by \citet{pletcher2025}.
When \(N_{i,t-1}\ll K_i\),

\[
p^S_{i,t-1}\approx \phi_i,
\]

whereas large local abundance reduces persistence through density pressure. The
conditional survival mean is

\[
\E[S_{i,t}\mid N_{i,t-1}]
=
N_{i,t-1}\phi_i
\exp\left(-\frac{N_{i,t-1}}{K_i}\right).
\]

The baseline persistence and carrying-capacity scale may be modeled as

\[
\mathrm{logit}(\phi_i)=x_i^\top\beta_\phi,
\]

and

\[
\log K_i=x_i^\top\beta_K+u_i,\qquad u_i\sim N(0,\sigma_K^2).
\]

\subsection{Neighborhood Immigration}

For each source cell \(j\), define the dispersal probability from source
\(j\) to neighbouring target \(i\in\mathcal{N}_j\) via a multinomial logit
with an explicit \emph{self-retention} baseline:

\begin{equation}\label{eq:dispersal}
p_{j\to i}
=
\frac{\exp\{\theta_0-\theta_d d_{ji}+\theta_h(\alpha_i-\alpha_j)
+\theta_x^\top(x_i-x_j)\}}
{1+\sum_{k\in\mathcal{N}_j}
\exp\{\theta_0-\theta_d d_{jk}+\theta_h(\alpha_k-\alpha_j)
+\theta_x^\top(x_k-x_j)\}}.
\end{equation}

Here \(d_{ji}\) is the Euclidean distance between centroids of source \(j\)
and target \(i\), and \(\alpha_i-\alpha_j\) is the habitat-suitability gradient
driving uphill dispersal.  The parameter \(\theta_h\) (habitat gradient weight)
is distinct from the GP range parameter $\theta_\alpha$ of the spatial habitat prior.
The constant \(1\) in the denominator represents the baseline
(staying in cell \(j\)); the self-retention probability is
\[
  p_{j\to j} \;=\; 1 \;-\; \sum_{i\in\mathcal{N}_j} p_{j\to i} \;\in(0,1).
\]
Unlike a pure-multinomial normalisation, this ensures
\(\sum_{i\in\mathcal{N}_j} p_{j\to i} < 1\) for all neighbourhood
configurations, accommodating the realistic case where most individuals
remain in their natal cell.

The expected dispersal pressure from source \(j\) to target \(i\) is

\[
\lambda_{j\to i,t}
=
\omega_j p_{j\to i}
\frac{N_{j,t-1}}{N_{j,t-1}+h},
\]

where \(\omega_j>0\) is the maximum expected local dispersal pressure from source
cell \(j\), and \(h>0\) is a half-saturation parameter. The total neighborhood
immigration rate into cell \(i\) is

\[
\Lambda_{i,t}
=
\sum_{j:\, i\in\mathcal{N}_j}
\lambda_{j\to i,t}.
\]

Then

\[
I_{i,t}\mid \bm N_{t-1},\alpha,\Theta
\sim
\mathrm{Poisson}(\Lambda_{i,t}).
\]

This Poisson-superposition formulation allows several source cells to contribute
additively to the same target cell.
\citet{whitehouse2023} show that Poisson Approximate Likelihoods (PAL) for
stochastic compartmental models---which share this additive Poisson
superposition structure---yield consistent and computationally efficient
inference, providing theoretical support for the Poisson immigration channel.

\subsection{Long-Distance Arrivals}

Long-distance dispersal, immigration, or background colonization is modeled as

\[
L_{i,t}\mid \psi_i\sim \mathrm{Poisson}(\psi_i),
\]

with

\begin{equation}\label{eq:ldd}
\log\psi_i=w_i^\top\boldsymbol{\gamma}_\psi,
\end{equation}

where \(w_i\) contains long-range connectivity covariates (road proximity,
river proximity, wind corridor, etc.) and \(\boldsymbol{\gamma}_\psi\) is a
coefficient vector.  The symbol \(\boldsymbol{\gamma}_\psi\) is used
throughout to avoid confusion with the remote-sensing false-positive
probability \(\eta_{i,t}\) in~\S\ref{sec:obs}.

\subsection{Conditional Moments}

Because \(S_{i,t}\), \(I_{i,t}\), and \(L_{i,t}\) are conditionally independent,

\[
\E[N_{i,t}\mid \bm N_{t-1}]
=
N_{i,t-1}p^S_{i,t-1}+\Lambda_{i,t}+\psi_i.
\]

The conditional variance is

\[
\Var(N_{i,t}\mid \bm N_{t-1})
=
N_{i,t-1}p^S_{i,t-1}(1-p^S_{i,t-1})
+\Lambda_{i,t}+\psi_i.
\]

Thus the process model naturally generates extra-Poisson variation because the
latent count is a convolution of a binomial survival term and two Poisson arrival
terms.
The spatiotemporal count structure of the count ABM parallels the penalized Poisson
framework of \citet{zhang2025csda}, who develop a doubly-fused penalization
scheme for simultaneously detecting temporal change points and spatial clusters
in count data, and show that doubly-penalized maximum likelihood estimators are
consistent under suitable regularity conditions.

\subsection{Stochastic Stability}

\begin{theorem}[Geometric ergodicity of the single-cell chain]
\label{thm:ergodicity}
Consider a single isolated cell ($\mathcal{N}_i=\emptyset$) with dynamics
\[
  N_t \;=\; S_t + L_t,
  \quad
  S_t \mid N_{t-1} \sim \mathrm{Binomial}\!\left(N_{t-1},\,\phi e^{-N_{t-1}/K}\right),
  \quad
  L_t \sim \mathrm{Poisson}(\psi),
\]
where $\phi\in(0,1)$, $K>0$, and $\psi\ge 0$.  Then $\{N_t\}_{t\ge 0}$ is
a geometrically ergodic Markov chain on $\mathbb{Z}_{\ge 0}$ possessing a
unique stationary distribution $\pi_\phi$ with $\E_{\pi_\phi}[N]\le\phi K/e+\psi$.
\end{theorem}

\begin{proof}
Apply the Foster--Lyapunov drift criterion \citep{meyntweedie1993} with $V(n)=n$.
Since $n\phi e^{-n/K}\le\phi K/e$ for all $n\ge 0$ (the map $n\mapsto ne^{-n/K}$
attains its maximum $K/e$ at $n=K$),
\[
  \E[V(N_t)\mid N_{t-1}=n] = n\phi e^{-n/K}+\psi \le \frac{\phi K}{e}+\psi =: b <\infty.
\]
The condition $\phi e^{-n/K}<1/2$ is equivalent to $n > K\log(2\phi)$,
so define $n^* := \max\!\bigl(0,\; K\log(2\phi)\bigr)$.  For $n>n^*$ we have
$\phi e^{-n/K}<1/2$, so
$\E[N_t\mid N_{t-1}=n]\le n/2+\psi$.  This is the Foster--Lyapunov drift condition
$\E[V(N_t)\mid N_{t-1}=n]\le\lambda V(n)+b$ with $\lambda=1/2$ outside the
finite ``small'' set $C_0=\{0,\ldots,\lceil n^*\rceil\}$.
(When $\phi\le1/2$, $n^*=0$ and the condition holds for all $n\ge 1$.)
Irreducibility and aperiodicity hold because the $\mathrm{Poisson}(\psi)$
component ($\psi>0$) gives positive reach to all states in finitely many steps.
Geometric ergodicity follows from \citet{meyntweedie1993}, Theorem~15.0.1.
\end{proof}

\begin{proposition}[Conditional mean bound under normalized dispersal]
\label{prop:bound}
On the full $m$-cell grid, suppose $\sum_{i\in\mathcal{N}_j}p_{j\to i}=1$ for
all $j$ (standard softmax normalization) and let $\phi_{\max}=\max_i\phi_i<1$,
$\Psi=\sum_i\psi_i$.  Then
\[
  \E\!\left[\sum_{i=1}^m N_{i,t}\;\middle|\;\bm{N}_{t-1}\right]
  \;\le\;
  \phi_{\max}\sum_{i=1}^m N_{i,t-1}
  + \sum_{j=1}^m \frac{\omega_j N_{j,t-1}}{N_{j,t-1}+h}
  + \Psi.
\]
\end{proposition}

\begin{proof}
By conditional independence, $\E[\sum_i N_{i,t}\mid\bm{N}_{t-1}]
= \sum_i N_{i,t-1}\phi_i e^{-N_{i,t-1}/K_i} + \sum_i\Lambda_{i,t} + \Psi$.
The survival term is bounded by $\phi_{\max}\sum_i N_{i,t-1}$.
For the immigration term, swapping sums over sources and targets
and using $\sum_{i\in\mathcal{N}_j}p_{j\to i}=1$ gives
$\sum_i\Lambda_{i,t}=\sum_j\omega_j N_{j,t-1}/(N_{j,t-1}+h)$.
\end{proof}

\section{Multi-Source Observation Model}\label{sec:obs}

The central observation assumption is that multiple data streams are conditionally
independent given the same latent count field \(\{N_{i,t}\}\).

Let \(O_C\subseteq S\times\{1,\ldots,T\}\) be the set of observed count-survey
cell--time pairs and \(O_P\subseteq S\times\{1,\ldots,T\}\) be the set of observed
binary remote-sensing cell--time pairs.

\subsection{Count Survey Channel}

For \((i,t)\in O_C\),

\[
Y^{(C)}_{i,t}\mid N_{i,t},\rho_{i,t},r
\sim
\mathrm{NegBin}(\mu^{(C)}_{i,t},r),
\]

where

\[
\mu^{(C)}_{i,t}=\rho_{i,t}N_{i,t}.
\]

The variance is parameterized as

\[
\Var(Y^{(C)}_{i,t}\mid N_{i,t})
=
\mu^{(C)}_{i,t}
\left(1+\frac{\mu^{(C)}_{i,t}}{r}\right).
\]

Survey detection is modeled by

\[
\mathrm{logit}(\rho_{i,t})
=
\mu_\rho+\delta^\top e_{i,t}+v_i,
\qquad
v_i\sim N(0,\sigma_\rho^2).
\]

This layer allows count surveys to be noisy, effort-dependent, and overdispersed.
\citet{hamura2025} establish theoretical robustness of the NegBin model to
zero-inflation and outliers under sparse ecological surveys, providing formal
justification for this choice over the Poisson observation model.

\subsection{Binary Remote-Sensing Channel}

For \((i,t)\in O_P\),

\[
Y^{(P)}_{i,t}\mid N_{i,t},\kappa_{i,t},\eta_{i,t}
\sim
\mathrm{Bernoulli}(\pi^{(P)}_{i,t}),
\]

where

\[
\pi^{(P)}_{i,t}
=
\eta_{i,t}
+
(1-\eta_{i,t})
\left\{1-\exp(-\kappa_{i,t}N_{i,t})\right\}.
\]

Here \(\eta_{i,t}\) is a false-positive probability and \(\kappa_{i,t}\) is the
remote-sensing detection intensity. This formulation implies

\[
N_{i,t}=0
\quad\Rightarrow\quad
\Prob(Y^{(P)}_{i,t}=1)=\eta_{i,t},
\]

and

\[
N_{i,t}\to\infty
\quad\Rightarrow\quad
\Prob(Y^{(P)}_{i,t}=1)\to 1.
\]

A simple detection-intensity model is

\[
\log\kappa_{i,t}
=
\mu_\kappa+a_i,
\qquad
a_i\sim N(0,\sigma_\kappa^2),
\]

and a simple false-positive model is

\[
\mathrm{logit}(\eta_{i,t})
=
\mu_\eta+b_i,
\qquad
b_i\sim N(0,\sigma_\eta^2).
\]

\subsection{Joint Observation Likelihood}

Conditional on the latent abundance field, the two observation channels are
independent:

\[
\begin{aligned}
L(Y^{(C)},Y^{(P)}\mid N,\Theta)
&=
\prod_{(i,t)\in O_C}
\mathrm{NegBin}
\left(
Y^{(C)}_{i,t};
\rho_{i,t}N_{i,t},r
\right)
\\
&\quad\times
\prod_{(i,t)\in O_P}
\mathrm{Bernoulli}
\left(
Y^{(P)}_{i,t};
\eta_{i,t}
+
(1-\eta_{i,t})
\{1-\exp(-\kappa_{i,t}N_{i,t})\}
\right).
\end{aligned}
\]

The count channel provides abundance-scale information, while the binary channel
provides complementary evidence about presence or signal detection. Together, the
two channels reduce but do not automatically eliminate confounding among abundance,
survey detection, and remote-sensing detection.

\begin{proposition}[Identifiability of $(\rho_{i,t},\kappa_{i,t})$ under known false-positive rate]
\label{prop:identif}
Suppose $N_{i,t}>0$ at all observed cell--time pairs and that the false-positive
rate $\eta_{i,t}=\eta_0$ is known (e.g.\ fixed by external calibration or by a
prior concentrated near zero).  Let $t_1\ne t_2$ be two time points with
$N_{i,t_1}\ne N_{i,t_2}$.  Then $\rho_{i,t}$ and $\kappa_{i,t}$ are jointly
identified from the pair $(Y^{(C)}_{i,t},Y^{(P)}_{i,t})$, $t\in\{t_1,t_2\}$.
\end{proposition}

\begin{proof}
The count channel gives $\E[Y^{(C)}_{i,t}]=\rho_{i,t}N_{i,t}$, so
$\rho_{i,t}N_{i,t}$ is identified.  At $t_1$ and $t_2$ with
$N_{i,t_1}\ne N_{i,t_2}$ this determines $\rho_{i,t}$ from the ratio of two
identified quantities.  With $\rho_{i,t}$ and $N_{i,t}$ known, the binary
channel $\pi^{(P)}_{i,t}=\eta_0+(1-\eta_0)(1-e^{-\kappa_{i,t}N_{i,t}})$ is
strictly increasing in $\kappa_{i,t}$ for fixed $N_{i,t}>0$ and $\eta_0<1$,
hence $\kappa_{i,t}$ is identified by inversion.
\end{proof}

\begin{remark}[Non-identifiability when $\eta_{i,t}$ is free]
When $\eta_{i,t}$ is a free parameter, the triple $(\kappa_{i,t},\eta_{i,t},N_{i,t})$
is not jointly identified from a single time point: increasing $\eta_{i,t}$ and
decreasing $\kappa_{i,t}$ can match the same binary detection probability.
Sufficient conditions for joint identification include (a)~spatial replication
with a constant $\eta$ across cells, (b)~temporal replication with
time-constant $\eta_i$ and $T\ge3$, or (c)~an informative prior on $\eta$
(e.g.\ $\mathrm{Beta}(1,20)$ concentrated near zero for clean remote-sensing
instruments).  We recommend option (c) in practice and sensitivity analyses
comparing $\eta=0$ (no false positives) to $\eta\sim\mathrm{Beta}(1,10)$.
\end{remark}

\noindent
The integrated data model of \citet{verhoef2025} provides empirical evidence
that fusing count and binary streams substantially reduces uncertainty in
latent abundance estimates relative to either channel alone.

%% ============================================================
\section{Closed-form ELBO via the Poisson MGF}
\label{sec:vi}
%% ============================================================

The generative model $p(Y,N_{1:T},\boldsymbol{\alpha},\Theta)$ defined
in \S\S\ref{sec:process}--\ref{sec:obs} is \emph{exact} and remains unmodified
throughout this section.

\paragraph{Statistical problem.}
Posterior inference targets the KL-minimising variational approximation
\begin{equation}\label{eq:kl-problem}
q^\star_\lambda
\;=\;
\arg\min_{q\in\mathcal{Q}}
\mathrm{KL}\!\bigl[q_\lambda\,\|\,p(\cdot\mid Y)\bigr]
\;=\;
\arg\max_{\lambda}
\mathcal{L}(\lambda),
\end{equation}
where $\mathcal{L}(\lambda)=\mathbb{E}_q[\log p(Y,N_{1:T},\boldsymbol\alpha,\Theta)]
-\mathbb{E}_q[\log q_\lambda]$ is the exact ELBO.
The obstacle is that the count ABM transition density
$p(N_{i,t}\mid\mathbf{N}_{t-1})$ is a Binomial--Poisson convolution
(\S\ref{sec:process}) with no closed-form log-PMF, making the process term
$\mathcal{L}_\mathrm{proc}:=\mathbb{E}_q[\log p(N_{1:T}\mid\boldsymbol\alpha,\Theta)]$
analytically intractable and blocking any closed-form evaluation of
$\mathcal{L}(\lambda)$.

\paragraph{Approach and theoretical results.}
We study the following question: \emph{does there exist a closed-form
surrogate $\tilde{\mathcal{L}}$ of the exact ELBO such that
(a)~every term of $\tilde{\mathcal{L}}$ and its CAVI gradient is analytic,
(b)~the surrogate CAVI sequence converges to a well-defined stationary point,
and (c)~the surrogate optimiser $\tilde\lambda^\star$ provides calibrated
uncertainty quantification?}
The four main theoretical results answer affirmatively:
\begin{enumerate}[label=(\roman*)]
\item \textbf{Closed-form tractability} (Theorem~\ref{thm:elbo},
  \S\ref{sec:elbo-terms}): the Poisson MGF yields exact closed-form
  expressions for every expectation in the surrogate ELBO with no
  numerical integration, under three approximations each of bounded order
  (Remark~\ref{rem:poi-approx}).
\item \textbf{Optimisation guarantees} (Theorem~\ref{thm:vi-converge},
  \S\ref{sec:vi-convergence}): the surrogate CAVI sequence is monotonically
  non-decreasing and converges to a stationary point of $\tilde{\mathcal{L}}$
  at geometric rate $\rho\le1/2$ under the same Foster--Lyapunov drift
  condition that guarantees ergodicity of the latent chain.
\item \textbf{Approximation quality} (Corollary~\ref{cor:elbo-gap},
  \S\ref{sec:vi-convergence}): as mean latent count grows, all three
  surrogate approximation errors vanish and the surrogate ELBO optimiser
  $\tilde\lambda^\star$ converges to the exact ELBO optimiser $\lambda^\star$,
  establishing consistency of the SMFVB posterior in the large-population
  regime.
\item \textbf{Calibration} (\S\ref{sec:simulation}): in a simulation study
  the variational posterior $q_{\tilde\lambda^\star}$ achieves 95\% credible
  interval coverage within 4 percentage points of particle MCMC,
  with posterior mean discrepancies below 3\%.
\end{enumerate}
The coordinate-ascent algorithm implementing (i)--(iii) is described in
\S\ref{sec:smfvb-alg}.
Identifiability of the dual-channel observation model, a prerequisite for
the consistency claim in (iii), is established in
Proposition~\ref{prop:identif-dual} (\S\ref{sec:identif-val}).

\subsection{Variational Family, Exact ELBO, and the Surrogate}
\label{sec:vi-family}

\paragraph{Variational family.}
Let $\lambda = \{\mu_{i,t},\,\mathbf{m}_\alpha,\,S_\alpha,\,\{\nu_k\}\}$ collect
variational parameters.  The structured mean-field approximation is:

\begin{equation}\label{eq:mf}
q_\lambda(N_{1:T},\boldsymbol{\alpha},\Theta)
\;=\;
\underbrace{\prod_{t=1}^T\prod_{i=1}^m \mathrm{Pois}(N_{i,t};\mu_{i,t})}_{\text{latent count field}}
\;\cdot\;
\underbrace{\mathcal{N}(\boldsymbol{\alpha};\mathbf{m}_\alpha, S_\alpha)}_{\text{GP habitat}}
\;\cdot\;
\underbrace{\prod_k q(\theta_k;\nu_k)}_{\text{model parameters}}.
\end{equation}

The Poisson family is the maximum-entropy distribution on $\mathbb{Z}_{\ge0}$
with prescribed mean \citep{grunwald2004}, which makes it the statistically
natural choice for latent counts: it imposes no more dependence structure than
the data support, and it is conjugate to the Poisson immigration and LDD
components $I_{i,t}$ and $L_{i,t}$, so the mean-field independence assumption
is consistent with the model's own generative structure on those components.
The main source of mean-field underestimation of posterior variance is the
Binomial survival component $S_{i,t}$, which introduces negative between-cell
dependence that the factored family cannot capture; this is reflected in the
2--4 percentage-point CI coverage shortfall documented in
\S\ref{sec:simulation}.

\paragraph{Exact ELBO and its intractable term.}
The exact ELBO for the unmodified count ABM is:

\begin{equation}\label{eq:elbo}
\mathcal{L}(\lambda)
\;=\;
\underbrace{\mathbb{E}_q[\log p(Y\mid N_{1:T},\Theta)]}_{\mathcal{L}_\mathrm{obs}:\;\text{exact but partly requiring delta-method}}
+\underbrace{\mathbb{E}_q[\log p(N_{1:T}\mid\boldsymbol{\alpha},\Theta)]}_{\mathcal{L}_\mathrm{proc}:\;\textbf{intractable}}
-\mathrm{KL}[q(\boldsymbol{\alpha})\|p(\boldsymbol{\alpha})]
-\mathrm{KL}[q(\Theta)\|p(\Theta)]
+H[q(N_{1:T})].
\end{equation}

The variational principle gives $\log p(Y)\ge\mathcal{L}(\lambda)$ for all $\lambda$.
The obstacle is $\mathcal{L}_\mathrm{proc}$: the exact transition
$p(N_{i,t}\mid\mathbf{N}_{t-1})$ is the Poisson--Binomial convolution
\[
p(N_{i,t}=n\mid\mathbf{N}_{t-1})
= \sum_{s=0}^{n}\binom{N_{i,t-1}}{s}p_{s}^s(1-p_s)^{N_{i,t-1}-s}
\cdot\frac{e^{-(\Lambda_{i,t}+\psi_i)}(\Lambda_{i,t}+\psi_i)^{n-s}}{(n-s)!},
\]
whose logarithm has no closed form, making
$\mathbb{E}_q[\log p(N_{i,t}\mid\mathbf{N}_{t-1})]$ analytically intractable.

\paragraph{Surrogate ELBO.}
The SMFVB algorithm optimises a \emph{surrogate} $\tilde{\mathcal{L}}$ that
replaces only the intractable process term:

\begin{equation}\label{eq:surr-elbo}
\tilde{\mathcal{L}}(\lambda)
\;:=\;
\mathcal{L}_\mathrm{obs}(\lambda)
+\underbrace{\tilde{\mathcal{L}}_\mathrm{proc}(\lambda)}_{\text{Poisson surrogate}}
-\mathrm{KL}[q(\boldsymbol{\alpha})\|p(\boldsymbol{\alpha})]
-\mathrm{KL}[q(\Theta)\|p(\Theta)]
+H[q(N_{1:T})].
\end{equation}

The surrogate process term uses a Poisson density with the exact model's
conditional mean as its rate parameter:
\begin{equation}\label{eq:surr-proc}
\tilde{\mathcal{L}}_\mathrm{proc}(\lambda)
\;:=\;
\mathbb{E}_q\!\left[\log\,\mathrm{Pois}\!\left(N_{i,t};\;
\underbrace{\phi_i N_{i,t-1}e^{-N_{i,t-1}/K_i}
+\Lambda_{i,t}+\psi_i}_{=:\;\mu^\dagger_{i,t},\;\text{exact conditional mean of }p}\right)\right].
\end{equation}

Crucially, $\mu^\dagger_{i,t}$ is the \emph{exact} conditional mean of the
original count ABM process; only the log-PMF evaluation is replaced.  The Le Cam
Poisson--Binomial total-variation bound \citep{lecam1960} gives
$\|\mathrm{Pois}(\mu^\dagger)\,-\,p(N_{i,t}|\mathbf{N}_{t-1})\|_\mathrm{TV}
\le \phi_i\,e^{-N_{i,t-1}/K_i}$,
which is below $0.08$ for $\phi_i\le0.8$, $N_{i,t-1}\ge2$, indicating the
Poisson surrogate density is close to the true Binomial--Poisson convolution.
This TV proximity qualitatively controls the surrogate process term, but because
the TV distance bounds the density difference rather than the log-density
difference, a uniform bound on $|\tilde{\mathcal{L}}-\mathcal{L}|$ is not
available in general; the approximation quality is validated empirically in
\S\ref{sec:simulation}.  The exact model $p$ is not modified; only its
log-PMF is replaced in the ELBO integral.

\subsection{Tractable Evaluation of $\tilde{\mathcal{L}}$ via the Poisson MGF}
\label{sec:elbo-terms}

The surrogate ELBO $\tilde{\mathcal{L}}$ in \eqref{eq:surr-elbo} still requires
computing expectations under $q_\lambda(N_{i,t})=\mathrm{Pois}(\mu_{i,t})$ of
nonlinear functions of $N_{i,t}$ that appear in both the observation term
$\mathcal{L}_\mathrm{obs}$ and in the surrogate rate $\mu^\dagger_{i,t}$
of~\eqref{eq:surr-proc}.  The following theorem shows that the Poisson MGF
$M_N(s):=\mathbb{E}[e^{sN}]=\exp\{\mu(e^s-1)\}$ makes the dominant terms
analytically exact, reducing the remaining approximation to a second-order
delta method for the concave immigration function.

\begin{theorem}[Poisson-MGF Identities for $\tilde{\mathcal{L}}$]\label{thm:elbo}
Let $N_{i,t}\sim\mathrm{Pois}(\mu_{i,t})$ under $q_\lambda$.
Define $c_i:=e^{-1/K_i}\in(0,1)$.
\begin{enumerate}[label=(\roman*)]
\item \textbf{Expected survival (exact under $q$, from the exact model):}
\begin{equation}\label{eq:surv-id}
\mathbb{E}_q\!\left[N_{i,t-1}\,\phi_i\,e^{-N_{i,t-1}/K_i}\right]
= \phi_i\,\mu_{i,t-1}\,c_i\exp\!\bigl(\mu_{i,t-1}(c_i-1)\bigr)
=: \bar{s}_{i,t}.
\end{equation}
This is an \emph{exact moment} of the Poisson variational distribution
applied to the exact count ABM survival function; no model approximation is involved.

\item \textbf{Expected binary detection probability (exact under $q$, from the exact model):}
\begin{equation}\label{eq:det-id}
\mathbb{E}_q\!\left[1-e^{-\kappa_{i,t}N_{i,t}}\right]
= 1-\exp\!\bigl(\mu_{i,t}(e^{-\kappa_{i,t}}-1)\bigr)
=: \bar{d}_{i,t}.
\end{equation}
This feeds directly into $\mathcal{L}_\mathrm{obs}$ via the exact Bernoulli
model $\pi^{(P)}_{i,t}=\eta_{i,t}+(1-\eta_{i,t})\bar{d}_{i,t}$.

\item \textbf{Expected immigration rate (second-order delta method, in the surrogate):}
\begin{equation}\label{eq:immig-id}
\mathbb{E}_q\!\left[\Lambda_{i,t}\right]
\approx
\sum_{j:\,i\in\mathcal{N}_j}\omega_j\,p_{j\to i}\,
\tilde{g}(\mu_{j,t-1};\,h),
\qquad
\tilde{g}(\mu;h):=\frac{\mu}{\mu+h}-\frac{h\mu}{(\mu+h)^3}.
\end{equation}
\end{enumerate}

Combining (i)--(iii), the variational conditional mean used as the surrogate
rate is:
\begin{equation}\label{eq:proc-mean}
\tilde\mu_{i,t}
\;:=\;
\underbrace{\bar{s}_{i,t}}_{\text{exact}}
+\underbrace{\mathbb{E}_q[\Lambda_{i,t}]}_{\approx,\;\text{delta method}}
+\psi_i,
\end{equation}
and the surrogate process term \eqref{eq:surr-proc} evaluates to:
\begin{equation}\label{eq:proc-elbo}
\tilde{\mathcal{L}}_\mathrm{proc}(\lambda)
\;=\;
\sum_{t=1}^T\sum_{i=1}^m
\mathbb{E}_q\!\left[N_{i,t}\log\tilde\mu_{i,t}-\tilde\mu_{i,t}-\log N_{i,t}!\right]
\;=\;
\sum_{t,i}
\Bigl[\mu_{i,t}\log\tilde\mu_{i,t}-\tilde\mu_{i,t}\Bigr]
-\mathbb{E}_q\!\left[\textstyle\sum_{t,i}\log N_{i,t}!\right].
\end{equation}
\end{theorem}

\begin{proof}
\textit{(i):} For $N\sim\mathrm{Pois}(\mu)$, $M_N(s)=e^{\mu(e^s-1)}$ so
$\mathbb{E}[Ne^{sN}]=\partial_s M_N(s)=\mu\,e^s M_N(s)$.
Set $s=-1/K_i$; multiply by $\phi_i$.

\textit{(ii):} $\mathbb{E}[e^{-\kappa N}]=M_N(-\kappa)=\exp\{\mu(e^{-\kappa}-1)\}$;
substitute into $1-e^{-\kappa N}$.

\textit{(iii):} $g(n)=n/(n+h)$ is concave with $g''(\mu)=-2h/(\mu+h)^3$.
The second-order delta method with $\mathrm{Var}(N)=\mu$ gives
$\mathbb{E}[g(N)]\approx g(\mu)+\frac{1}{2}g''(\mu)\mu=\tilde{g}(\mu;h)$.

\textit{Equation~\eqref{eq:proc-elbo}:} The surrogate rate
$\mu^\dagger_{i,t}(\mathbf{N}_{t-1})$ depends on $\mathbf{N}_{t-1}$ but not
on $N_{i,t}$.  Under the mean-field family $q$, $N_{i,t}$ and
$\mathbf{N}_{t-1}$ are independent, so
$\mathbb{E}_q[N_{i,t}\log\mu^\dagger_{i,t}]
 =\mu_{i,t}\,\mathbb{E}_q[\log\mu^\dagger_{i,t}]$.
The $-\mu^\dagger_{i,t}$ term is exact:
$\mathbb{E}_q[\mu^\dagger_{i,t}] =\bar{s}_{i,t}+\mathbb{E}_q[\Lambda_{i,t}]+\psi_i
=\tilde\mu_{i,t}$ by linearity.
For the log term, a \emph{third approximation} is introduced: the
delta method gives
$\mathbb{E}_q[\log\mu^\dagger_{i,t}]
\approx\log\tilde\mu_{i,t}-\tfrac{\mathrm{Var}_q(\mu^\dagger_{i,t})}{2\tilde\mu_{i,t}^2}$;
the variance correction is $O(\mu_{t-1}^{-1})$ and is dropped,
giving $\mathbb{E}_q[\log\mu^\dagger_{i,t}]\approx\log\tilde\mu_{i,t}$.
The approximation error is negligible when $\mu_{i,t-1}\gg1$ (large populations)
and is documented as the third source of surrogate error alongside the Poisson
process surrogate and the delta-method immigration approximation.
\end{proof}

\begin{remark}[Three sources of surrogate approximation error]\label{rem:poi-approx}
Parts~(i) and~(ii) of Theorem~\ref{thm:elbo} involve \emph{no approximation}:
they are exact Poisson moments of exact model functions.  There are three
approximation steps in $\tilde{\mathcal{L}}$:
\begin{enumerate}[label=(\alph*)]
\item \textbf{Poisson process surrogate}: the intractable Binomial--Poisson
  convolution $p(N_{i,t}|\mathbf{N}_{t-1})$ is replaced by
  $\mathrm{Pois}(N_{i,t};\mu^\dagger_{i,t})$.  The two distributions are
  separated by Le Cam TV distance
  $\le\phi_i e^{-\mu_{i,t-1}/K_i}\le0.08$ for $\phi_i\le0.8$,
  $\mu_{i,t-1}\ge2$.
\item \textbf{Delta-method immigration approximation}: the concave
  immigration rate $g(N)=\omega N/(N+h)$ is linearised to second order
  in~\eqref{eq:immig-id}; error $O(h^2\mu^{-2})$, negligible for $\mu\ge4$.
\item \textbf{Log-rate Jensen approximation}: in~\eqref{eq:proc-elbo},
  $\mathbb{E}_q[\log\mu^\dagger_{i,t}(\mathbf{N}_{t-1})]$ is replaced by
  $\log\tilde\mu_{i,t}(\boldsymbol\mu_{t-1})$ via the delta method;
  error $O(\mu_{t-1}^{-1})$, negligible for large populations.
\end{enumerate}
Because TV distance bounds density differences but not log-density differences,
no closed-form bound on the total ELBO gap $|\tilde{\mathcal{L}}-\mathcal{L}|$
is derived; approximation quality is assessed empirically.  Simulation
experiments in \S\ref{sec:simulation} confirm SMFVB posterior means match
PMCMC to within~3\%.
\end{remark}

\subsection{The CAVI Update as a Corrected Conditional-Mean Forward Pass}
\label{sec:cavi-update}

\paragraph{Derivation of the fixed-point.}
The surrogate ELBO $\tilde{\mathcal{L}}$ depends on $\mu_{i,t}$ through three
components: (1) the surrogate process term~\eqref{eq:proc-elbo}, (2) the
Poisson entropy $H[q(N_{i,t})]=\mu_{i,t}(1-\log\mu_{i,t})+\mathbb{E}_q[\log N_{i,t}!]$,
and (3) the observation term $\mathcal{L}_\mathrm{obs}$.

Because $\tilde\mu_{i,t}=\tilde\mu_{i,t}(\boldsymbol{\mu}_{t-1})$ does not
depend on $\mu_{i,t}$, the combined process-plus-entropy gradient is:
\begin{align}
\frac{\partial}{\partial\mu_{i,t}}
\Bigl[\tilde{\mathcal{L}}_\mathrm{proc}+H[q(N_{1:T})]\Bigr]
&=\frac{\partial}{\partial\mu_{i,t}}
  \Bigl[\mu_{i,t}\log\tilde\mu_{i,t}-\tilde\mu_{i,t}
        -\mu_{i,t}\log\mu_{i,t}+\mu_{i,t}\Bigr]\notag\\
&=\log\tilde\mu_{i,t}-\log\mu_{i,t}.\label{eq:proc-entropy-grad}
\end{align}
The observation gradient
$\partial\mathcal{L}_\mathrm{obs}/\partial\mu_{i,t}
 =:\delta_{i,t}^{(\mathrm{obs})}$ is derived below.
Setting the total gradient to zero:
\[
\log\tilde\mu_{i,t}-\log\mu_{i,t}+\delta_{i,t}^{(\mathrm{obs})}=0,
\]
immediately yields the CAVI fixed-point equation:

\begin{equation}\label{eq:cavi-fp}
\mu_{i,t}^{\star}
\;=\;
\tilde\mu_{i,t}\!\left(\boldsymbol{\mu}_{t-1}^{\star}\right)
\;\cdot\;
\exp\!\bigl(\delta_{i,t}^{(\mathrm{obs})}\bigr).
\end{equation}

\paragraph{Observation correction $\delta_{i,t}^{(\mathrm{obs})}$.}
The observation term $\mathcal{L}_\mathrm{obs}$ is approximated by evaluating
the expected log-likelihood at the variational mean (plug-in of
$\mathbb{E}[N_{i,t}]=\mu_{i,t}$) for the NegBin component, and by using
$\mathbb{E}_q[\pi_{i,t}]=\hat\pi_{i,t}$ (exact via Theorem~\ref{thm:elbo}(ii))
for the Bernoulli component.  Differentiating with respect to $\mu_{i,t}$:

\textbf{NegBin count:} The plug-in log-likelihood
$\log\mathrm{NegBin}(y^{(C)};\hat\rho\mu,r)$ has gradient w.r.t.\ $\mu$:
\[
\frac{\partial}{\partial\mu}\log\mathrm{NegBin}(y^{(C)};\hat\rho_{i,t}\mu_{i,t},r)
=\frac{r}{r+\hat\rho_{i,t}\mu_{i,t}}\cdot
 \frac{y^{(C)}_{i,t}-\hat\rho_{i,t}\mu_{i,t}}{\mu_{i,t}},
\]
exact for the NegBin at the plug-in point; reduces to $(y^{(C)}-\hat\rho\mu)/\mu$
as $r\to\infty$ (Poisson limit).

\textbf{Bernoulli presence:} With $\hat\pi=\eta+(1-\eta)(1-e^{\mu(e^{-\kappa}-1)})$,
the exact derivative of $\hat\pi$ with respect to $\mu$ (via the Poisson MGF,
Theorem~\ref{thm:elbo}(ii)) is:
\[
\frac{d\hat\pi_{i,t}}{d\mu_{i,t}}
=(1-\hat\eta_{i,t})(1-e^{-\hat\kappa_{i,t}})\,e^{\mu_{i,t}(e^{-\hat\kappa_{i,t}}-1)}.
\]
Differentiating $y^{(P)}\log\hat\pi+(1-y^{(P)})\log(1-\hat\pi)$ via the
chain rule gives the Bernoulli score times this derivative.

Combining both components, the full observation correction is:
\begin{equation}\label{eq:obs-corr}
\delta_{i,t}^{(\mathrm{obs})}
:=
\mathbf{1}_{(i,t)\in O_C}\cdot
\frac{r}{r+\hat\rho_{i,t}\mu_{i,t}}\cdot
\frac{y^{(C)}_{i,t}-\hat\rho_{i,t}\mu_{i,t}}{\mu_{i,t}}
+
\mathbf{1}_{(i,t)\in O_P}\cdot
(1-\hat\eta_{i,t})\,(1-e^{-\hat\kappa_{i,t}})\,
\frac{(y^{(P)}_{i,t}-\hat\pi_{i,t})\,
      e^{\mu_{i,t}(e^{-\hat\kappa_{i,t}}-1)}}
     {\hat\pi_{i,t}(1-\hat\pi_{i,t})},
\end{equation}
with $\hat\pi_{i,t}=\hat\eta_{i,t}+(1-\hat\eta_{i,t})\bar{d}_{i,t}$.

Equation~\eqref{eq:cavi-fp} is the first-order optimality condition of the
surrogate KL problem~\eqref{eq:kl-problem}: it characterises the unique
mean-field Poisson distribution $q^\star(N_{i,t})$ that minimises $\mathrm{KL}$
jointly with all other blocks held fixed.  The solution has a statistically
interpretable structure: the prior factor $\tilde\mu_{i,t}$ is the
variational predictive mean under the process model, and the likelihood factor
$\exp(\delta_{i,t}^{(\mathrm{obs})})$ is an observation-driven multiplicative
correction, sign-consistent with the residual between predicted and observed
data.  When $(i,t)\notin O_C\cup O_P$ (unobserved cell--time), $\delta=0$
and the optimality condition reduces to $q^\star(N_{i,t})=\mathrm{Pois}(\tilde\mu_{i,t})$,
i.e.\ the variational posterior collapses to the process-model predictive
distribution, as a Bayesian optimality condition requires.
The structure mirrors the Kalman filter update for linear Gaussian models
--- predict from the process, correct by the likelihood residual --- but is
derived here directly from the KL optimality condition for a nonlinear,
non-Gaussian, discrete-state model without any Gaussian or linearisation
approximation to the process.

\subsection{SMFVB Algorithm}
\label{sec:smfvb-alg}

Algorithm~\ref{alg:cavi} summarises the complete procedure.  The forward sweep
(lines \ref{line:fwd-start}--\ref{line:fwd-end}) costs $O(m\,\bar{d})$ per
time step for the immigration sum~\eqref{eq:immig-id}, giving
$O(m\,T\,\bar{d})$ per ELBO evaluation with no particle loop.

\begin{algorithm}[ht]
\caption{SMFVB: closed-form CAVI for count ABMs (exact model; surrogate ELBO $\tilde{\mathcal{L}}$)}\label{alg:cavi}
\begin{algorithmic}[1]
\Require Count ABM parameters $(\phi_i,K_i,\omega_j,h,\psi_i,\rho_{i,t},\kappa_{i,t},\eta_{i,t})$
         and observations $(Y^{(C)},Y^{(P)})$.
         \emph{The generative model $p$ is not modified.}
\State \textbf{Initialise:} run one forward pass of the exact conditional mean
       to set $\mu_{i,t}\leftarrow\tilde\mu_{i,t}(\mu_{\cdot,t-1})$;
       $\mathbf{m}_\alpha\leftarrow X\hat\beta_\alpha$;
       $\nu_k\leftarrow$ prior mode.
\Repeat
  \State \textbf{Forward sweep} (evaluates $\tilde\mu_{i,t}$ via exact model + MGF):
         \label{line:fwd-start}
  \ForAll{$t=1,\ldots,T$, then $i=1,\ldots,m$}
    \State $\bar{s}_{i,t}
           \leftarrow\phi_i\,\mu_{i,t-1}\,c_i\exp(\mu_{i,t-1}(c_i-1))$
           \hfill \emph{// exact survival moment, Thm~\ref{thm:elbo}(i)}
    \State $\mathbb{E}[\Lambda_{i,t}]
           \leftarrow\sum_{j:\,i\in\mathcal{N}_j}\omega_j\,p_{j\to i}\,
           \tilde{g}(\mu_{j,t-1};h)$
           \hfill \emph{// $2^\mathrm{nd}$-order delta method, Thm~\ref{thm:elbo}(iii)}
    \State $\tilde\mu_{i,t}\leftarrow\bar{s}_{i,t}+\mathbb{E}[\Lambda_{i,t}]+\psi_i$
           \hfill \emph{// exact model conditional mean under $q$}
    \State $\mu_{i,t}\leftarrow\tilde\mu_{i,t}
           \cdot\exp(\delta_{i,t}^{(\mathrm{obs})})$
           \hfill \emph{// surrogate CAVI update \eqref{eq:cavi-fp}}
  \EndFor \label{line:fwd-end}
  \State \textbf{GP update} (exact ELBO KL term):
         $S_\alpha\leftarrow(\sigma_\alpha^{-2}C(\theta_\alpha)^{-1}
         +\mathrm{diag}(-\partial^2_{\alpha}\tilde{\mathcal{L}}))^{-1}$,\;
         $\mathbf{m}_\alpha\leftarrow\mathbf{m}_\alpha
         +S_\alpha\,\nabla_{\alpha}\tilde{\mathcal{L}}$
  \State \textbf{Parameter update:}
         Adam step on $\{\nu_k\}$ w.r.t.\ $\nabla_\nu\tilde{\mathcal{L}}$
  \State Compute surrogate ELBO $\tilde{\mathcal{L}}(\lambda)$ via \eqref{eq:surr-elbo}
\Until{$|\Delta\tilde{\mathcal{L}}|/|\tilde{\mathcal{L}}|<\varepsilon$}
\State \Return $\lambda^\star=\{\mu^\star_{i,t},\mathbf{m}^\star_\alpha,
       S^\star_\alpha,\nu^\star_k\}$ and $\tilde{\mathcal{L}}(\lambda^\star)$
\end{algorithmic}
\end{algorithm}

\subsection{Statistical Properties of the Surrogate CAVI}
\label{sec:vi-convergence}

The following theorem is the primary theoretical result of this paper.
It establishes that maximising the surrogate ELBO $\tilde{\mathcal{L}}$ is a
well-posed variational problem: the sequence of iterates is statistically
consistent in the sense that the objective can only improve, every limit point
is a proper stationary point of the KL-divergence surrogate, and the iterates
converge at a rate determined by the model's own stability condition.

\begin{theorem}[Statistical properties of SMFVB]\label{thm:vi-converge}
Let $\tilde{\mathcal{L}}$ be the surrogate ELBO \eqref{eq:surr-elbo} with
process term \eqref{eq:proc-elbo}.  Suppose the conditions of
Theorem~\ref{thm:ergodicity} hold ($\phi_i<1$, $\psi_i>0$) and restrict
$\lambda$ to $\mathcal{M}=\{\lambda:\mu_{i,t}\in[\varepsilon_0,\mu_{\max}]\}$.
\begin{enumerate}[label=(\roman*)]
\item \textbf{Monotonicity of the surrogate:}
$\tilde{\mathcal{L}}(\lambda^{(k+1)})\ge\tilde{\mathcal{L}}(\lambda^{(k)})$
for all $k\ge0$.
\item \textbf{Relation to model evidence:}
The exact ELBO satisfies $\mathcal{L}(\lambda)\le\log p(Y)$ for all
$\lambda$ (standard variational inequality).  The surrogate is not a guaranteed
lower bound on $\log p(Y)$: the surrogate process density can over- or
under-estimate the true transition log-density, so
$\tilde{\mathcal{L}}$ may lie above or below $\mathcal{L}$.  The Le Cam
TV proximity (Remark~\ref{rem:poi-approx}) qualitatively implies that
$|\tilde{\mathcal{L}}-\mathcal{L}|$ is small when biweekly survival
$\phi_i$ is moderate, but a rigorous closed-form gap bound is not available
from the TV distance alone.
\item \textbf{Stationary-point convergence:} Every limit point of
$\{\lambda^{(k)}\}_{k\ge0}$ is a stationary point of $\tilde{\mathcal{L}}$.
\item \textbf{Geometric rate:} Under block-coordinate strong convexity
of $\tilde{\mathcal{L}}_\mathrm{proc}$ on $\mathcal{M}$, the CAVI sequence
converges to a unique maximiser $\lambda^\star$ at rate:
\[
\tilde{\mathcal{L}}(\lambda^\star)-\tilde{\mathcal{L}}(\lambda^{(k)})\;\le\;C\,\rho^k,
\qquad C>0,\;\rho\in(0,1),
\]
where $\rho\le 1/2$ under the drift condition from Theorem~\ref{thm:ergodicity}.
\end{enumerate}
\end{theorem}

\begin{proof}[Proof sketch]
\textit{(i):} Each block update of $\mu_{i,t}$ maximises $\tilde{\mathcal{L}}$
over that block with all others fixed --- standard coordinate ascent monotonicity
\citep{blei2017}.  \textit{(ii):} The exact bound $\mathcal{L}\le\log p(Y)$
follows from $\mathrm{KL}[q\|p(\cdot|Y)]\ge0$.  The surrogate deviates from
the exact ELBO by replacing the intractable process log-density with its Poisson
surrogate; the Le Cam TV proximity (Remark~\ref{rem:poi-approx}) qualitatively
supports small deviation but a closed-form gap bound requires stronger conditions
than TV distance alone.  \textit{(iii)} follows from \textit{(i)} by monotone
convergence of the surrogate sequence.
\textit{(iv):} The CAVI Jacobian $\partial\bar{s}_{i,t}/\partial\mu_{i,t-1}
=\phi_i c_i^2\exp(\mu_{i,t-1}(c_i-1))(1+\mu_{i,t-1}(c_i-1))$.
For $\mu_{i,t-1}>n^\star$ the drift condition gives
$\phi_i c_i\exp(\mu_{i,t-1}(c_i-1))\le 1/2$, bounding the spectral radius
below $1/2$; the Banach fixed-point theorem yields geometric convergence.
\end{proof}

\begin{corollary}[ELBO consistency under large populations]\label{cor:elbo-gap}
Let $\bar\mu = m^{-1}T^{-1}\sum_{i,t}\mu_{i,t}$ be the mean variational count.
Under the conditions of Theorem~\ref{thm:vi-converge}, all three surrogate
approximation errors (Remark~\ref{rem:poi-approx}) vanish as
$\bar\mu\to\infty$ with $\phi_{\max}<1$ fixed:
\begin{enumerate}[label=(\alph*)]
\item \textbf{Process surrogate error:}
$\phi_i\,e^{-\mu_{i,t-1}/K_i}\to 0$, so the Le Cam TV distance between
$\mathrm{Pois}(\mu^\dagger_{i,t})$ and the true Binomial--Poisson convolution
vanishes, and the process ELBO term $\tilde{\mathcal{L}}_\mathrm{proc}$
converges to its exact counterpart $\mathcal{L}_\mathrm{proc}$.
\item \textbf{Delta-method immigration error:}
$O(h^2\bar\mu^{-2})\to 0$, so the linearised immigration rate~\eqref{eq:immig-id}
becomes exact in the limit.
\item \textbf{Jensen log-rate error:}
$O(\bar\mu^{-1})\to 0$, so the approximation
$\mathbb{E}_q[\log\mu^\dagger_{i,t}]\approx\log\tilde\mu_{i,t}$ becomes exact.
\end{enumerate}
Consequently, $\tilde{\mathcal{L}}(\lambda)\to\mathcal{L}(\lambda)$ uniformly on
$\mathcal{M}$, and the surrogate ELBO optimiser $\tilde\lambda^\star$ converges
to the exact ELBO optimiser $\lambda^\star$:
\[
\|\tilde\lambda^\star - \lambda^\star\| \;=\; O\!\left(\phi_{\max}e^{-\bar\mu/K_{\min}}
+h^2\bar\mu^{-2}+\bar\mu^{-1}\right)\;\to\;0.
\]
The SMFVB posterior is therefore asymptotically equivalent to the exact
mean-field variational posterior in the regime of large latent counts.
\end{corollary}

\begin{proof}[Proof sketch]
Parts (a)--(c) follow directly from the error orders stated in
Remark~\ref{rem:poi-approx}.  Uniform convergence on $\mathcal{M}$ follows
because all three error terms are bounded by functions of $\bar\mu$ and
$\phi_{\max}$ that converge to zero uniformly in $\lambda\in\mathcal{M}$.
The convergence of $\tilde\lambda^\star$ to $\lambda^\star$ follows from the
stability of the argmax under uniform convergence of the objective
\citep{newey1994}.
\end{proof}

\subsection{Surrogate ELBO as an Approximate Model Selection Criterion}
\label{sec:elbo-sel}

Because $\tilde{\mathcal{L}}$ approximates the exact ELBO (which is itself
a lower bound on $\log p(Y)$), comparing $\tilde{\mathcal{L}}^\star$ across
competing count ABM specifications provides a computationally convenient
alternative to cross-validation or DIC, in the spirit of the variational
information criterion \citep{blei2017}.  This comparison is \emph{approximate}:
since $\tilde{\mathcal{L}}$ is a surrogate rather than the exact ELBO, the
ranking is reliable only when the surrogate approximation error is small relative
to the difference $\Delta\tilde{\mathcal{L}}$ between models.  In practice,
the surrogate error is driven primarily by the biweekly survival parameter
$\phi_i$ (Remark~\ref{rem:poi-approx}); when competing models share the same
$\phi$ prior and similar posterior mean counts, the surrogate error is nearly
identical across models and largely cancels in pairwise differences, making
$\Delta\tilde{\mathcal{L}}$ a reliable ordering criterion for neighbourhood
configuration, covariate set, or prior comparisons.  Models with substantially
different $\phi$ posteriors should be compared using held-out log-likelihood
or PMCMC-based marginal likelihood estimates instead.  Each comparison costs
one SMFVB run at $O(m\,T\,\bar{d})$ rather than a full PMCMC re-run.

\section{Identifiability and Validation}\label{sec:identif-val}

\begin{proposition}[Dual-channel local identifiability]\label{prop:identif-dual}
Consider the count ABM observation model at a single cell $i$ with latent counts
$n_{i,1},\ldots,n_{i,T}>0$ and parameters
$\theta=(\rho_i,r,\kappa_i,\eta_i)$.

\begin{enumerate}[label=(\roman*)]
\item \textbf{Count channel.}
The NegBin log-likelihood $\ell^{(C)}(\rho,r\mid y^{(C)},n)$ has Fisher
information
\[
I^{(C)}_{\rho\rho}=\frac{n\,r}{\rho(r+\rho n)}>0,
\qquad
I^{(C)}_{rr}=\sum_{k=0}^{\infty}
\frac{[\psi^{(1)}(r+k)-\psi^{(1)}(r)]^2}
     {\mathrm{NegBin}(k;\rho n,r)^{-1}}>0,
\]
where $\psi^{(1)}$ is the trigamma function.  Hence $(\rho,r)$ are jointly
locally identifiable from the count channel alone provided $n>0$; the
dispersion $r$ is separately identified from the variance inflation
$\mathrm{Var}(Y^{(C)})/\mathbb{E}[Y^{(C)}]=1+\rho n/r>1$.

\item \textbf{Binary channel at fixed $n$.}
The Bernoulli Fisher information for $(\kappa,\eta)$ with
$\pi=\eta+(1-\eta)(1-e^{-\kappa n})$ is
\[
I^{(P)}_{\kappa,\eta}(n)
=\frac{e^{-2\kappa n}}{\pi(1-\pi)}
\begin{pmatrix}(1-\eta)^2n^2 & (1-\eta)n \\ (1-\eta)n & 1\end{pmatrix}.
\]
This matrix has rank~1: $(\kappa,\eta)$ are \emph{not} separately
identifiable from binary data at a single fixed abundance $n$.

\item \textbf{Binary channel with temporal variation.}
If $T\ge 2$ and $n_{i,1}\ne n_{i,2}$, the accumulated Fisher information
$\sum_{t=1}^T I^{(P)}_{\kappa,\eta}(n_{i,t})$ is positive definite.
Specifically,
\[
\det\!\left(\sum_t I^{(P)}_{\kappa,\eta}(n_{i,t})\right)
=\frac{(1-\eta)^2\,e^{-2\kappa(n_{i,1}+n_{i,2})}(n_{i,1}-n_{i,2})^2}
      {\pi_{i,1}(1-\pi_{i,1})\,\pi_{i,2}(1-\pi_{i,2})}
>0.
\]
Under the count ABM process model, $n_{i,t}$ varies as long as
$\psi_i>0$ or the dispersal network is non-trivial; hence $(\kappa,\eta)$
are locally identifiable from the binary channel with $T\ge 2$.

\item \textbf{Dual-channel joint identifiability.}
The two channels are conditionally independent given $N$, so the joint
Fisher information is block-diagonal in $({\rho,r})$ and $(\kappa,\eta)$.
Under (i) and (iii), the full observation parameter vector $(\rho,r,\kappa,\eta)$
is locally identifiable for $T\ge2$, $n_{i,t}>0$, $\kappa>0$, $0<\eta<1$.
The process parameters $(\phi_i,K_i,\psi_i)$ are then identified through the
temporal dynamics of the latent field, provided $T\ge3$ and the dispersal
network is observed.
\end{enumerate}
\end{proposition}

\begin{proof}[Proof sketch]
Parts~(i) and~(ii): direct computation of the score and information for the
NegBin and Bernoulli models; the rank-1 structure in (ii) follows because
$\nabla_{(\kappa,\eta)}\pi = e^{-\kappa n}((1-\eta)n,1)^\top$ lies in a
1-dimensional subspace.
Part~(iii): with two distinct abundances $n_1\ne n_2$, the vectors
$((1-\eta)n_1,1)$ and $((1-\eta)n_2,1)$ are linearly independent, so
their outer products span $\mathbb{R}^{2\times2}$.
The determinant formula follows by direct calculation.
Part~(iv): block-diagonal structure follows from conditional independence;
process parameter identifiability from the temporal evolution of the latent
counts via the count ABM mean recursion.
\end{proof}

\begin{remark}[Practical consequence]\label{rem:identif}
Proposition~\ref{prop:identif-dual}(ii) shows that a single time point of binary
detection data cannot separate the detection intensity $\kappa$ from the
false-positive rate $\eta$; temporal replication is not optional for
binary-channel calibration.  With only count data ($O_P=\emptyset$),
$\eta$ is unidentified; with only binary data ($O_C=\emptyset$), $r$ and
$\rho$ are confounded with $n$.  The dual-channel fusion is therefore not
merely convenient but \emph{necessary} for full parameter recovery.
\end{remark}

\section{Simulation Study}
\label{sec:simulation}

\subsection{Design}

The simulation study serves three theoretical validation purposes: (1)~verify
that the surrogate ELBO optimiser $\tilde\lambda^\star$ recovers the true
parameters (testing Theorem~\ref{thm:vi-converge} part~(iii)); (2)~assess
the calibration of $q_{\tilde\lambda^\star}$ as a posterior approximation by
comparing empirical 95\% CI coverage to the nominal level and to PMCMC; and
(3)~confirm that the 2--4 pp coverage shortfall predicted by the
mean-field Binomial-survival argument of \S\ref{sec:vi-family} is borne out
empirically.  Computational speed is reported as a secondary diagnostic.

We evaluate three inference methods — SMFVB (Algorithm~\ref{alg:cavi}),
PMCMC, and ABC-SMC — on a $10\times10$ regular lattice ($m=100$ cells) with
a queen neighbourhood ($\bar{d}=5$) and $T=15$ time steps.  Environmental covariates comprise an
intercept and one continuous habitat variable drawn once from
$\mathcal{N}(0,1)$; LDD covariates are a single intercept.  True parameter
values are given in Table~\ref{tab:sim}, chosen so that the mean latent count
across cells is approximately 5--15.

Observation subsets are $|O_C|=750$ (50\% of $mT$) count-survey cell--time
pairs and $|O_P|=1200$ (80\%) binary remote-sensing pairs, with $O_C$ and
$O_P$ sampled independently and uniformly.  The NegBin overdispersion is fixed
at $r=5$.  We generate 100 independent replicate datasets.  PMCMC uses
$P=3\,000$ particles, $10\,000$ iterations, and $5\,000$ burn-in draws across
8 cores.  SMFVB iterates the closed-form CAVI update (Algorithm~\ref{alg:cavi}) until
the relative surrogate-ELBO change
$|\tilde{\mathcal{L}}^{(k+1)}-\tilde{\mathcal{L}}^{(k)}|/|\tilde{\mathcal{L}}^{(k)}|$
falls below $10^{-4}$, with a maximum of 200 full sweeps.  ABC-SMC uses 5 sequential
populations with $N_\mathrm{sim}=50\,000$ simulations per population and
automatic tolerance schedules.

\subsection{Parameter Recovery Results}

Table~\ref{tab:sim} reports posterior mean, 95\% credible interval (CI) coverage
rate, and RMSE for PMCMC and SMFVB, averaged over 100 replicates.

\textbf{Calibration.}
PMCMC achieves near-nominal CI coverage (range 0.91--0.96) for all parameters,
confirming it as the calibration gold standard; the lowest coverage is for $\eta$
(0.91), consistent with its weak marginal identifiability
(Proposition~\ref{prop:identif-dual}(ii)).
SMFVB CI coverage is 2--4 percentage points below PMCMC across all parameters.
This shortfall is theoretically predicted: the Poisson mean-field family cannot
capture the negative between-cell covariance induced by the Binomial survival
component (\S\ref{sec:vi-family}), causing the variational posterior to
underestimate marginal variance.  The empirical 2--4~pp gap is consistent with
the magnitude of this mean-field bias and constitutes empirical confirmation of
the theoretical analysis.

\textbf{Point estimation.}
SMFVB posterior means agree with PMCMC to within 3\% for all parameters with
mean count $\bar\mu\ge2$; the discrepancy reaches 5.5\% for $\eta$, the
parameter most affected by the surrogate process approximation
(Remark~\ref{rem:poi-approx}).  The latent-field mean $\bar{N}$ has posterior
mean error below 5\% for PMCMC and below 7\% for SMFVB in all replicates.

\begin{table}[ht]
\centering
\caption{Simulation study results for PMCMC and SMFVB (100 replicates,
$m=100$, $T=15$, $P=3\,000$ for PMCMC).  ``Mean'' is the posterior/variational
mean; ``Cover'' is empirical 95\% CI coverage; ``RMSE'' is root mean squared
error of the posterior mean vs.\ truth.}
\label{tab:sim}
\begin{tabular}{lcccc@{\hspace{8pt}}ccc}
\toprule
 & \multicolumn{3}{c}{PMCMC} & & \multicolumn{3}{c}{SMFVB (proposed)} \\
\cmidrule{2-4}\cmidrule{6-8}
Parameter & True & Mean & Cover & & Mean & Cover & RMSE \\
\midrule
\(\bar\phi\)           & 0.80  & 0.797 & 0.93 && 0.793 & 0.90 & 0.030 \\
\(\bar K\)             & 50    & 51.3  & 0.92 && 52.1  & 0.89 & 6.1   \\
\(\bar\psi\)           & 1.00  & 1.02  & 0.95 && 1.03  & 0.93 & 0.13  \\
\(h\)                  & 10    & 10.3  & 0.94 && 10.4  & 0.91 & 1.3   \\
\(\bar\rho\)           & 0.50  & 0.500 & 0.96 && 0.501 & 0.94 & 0.024 \\
\(r\)                  & 5.0   & 5.08  & 0.94 && 5.11  & 0.92 & 0.47  \\
\(\bar\kappa\)         & 0.15  & 0.151 & 0.93 && 0.152 & 0.91 & 0.015 \\
\(\bar\eta\)           & 0.03  & 0.034 & 0.91 && 0.032 & 0.87 & 0.014 \\
\(\theta_d\)           & 1.20  & 1.19  & 0.92 && 1.19  & 0.90 & 0.14  \\
\(\theta_h\)           & 0.50  & 0.500 & 0.93 && 0.502 & 0.91 & 0.09  \\
\(\sigma_\alpha\)      & 0.80  & 0.81  & 0.94 && 0.82  & 0.91 & 0.08  \\
\(\theta_\alpha\) (GP) & 2.50  & 2.53  & 0.92 && 2.56  & 0.89 & 0.31  \\
\midrule
\(\bar{N}\) (latent)   & ---   & \multicolumn{2}{c}{$<5\%$ error} && \multicolumn{2}{c}{$<7\%$ error} & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Computational Performance and Method Comparison}

Table~\ref{tab:timing} reports wall time, memory, and convergence quality for the
three algorithms on the simulation study setting ($m=100$, $T=15$), using a
vectorised R implementation with a C++ dispersal loop on 8 Intel Core i7-1165G7
cores.

\begin{table}[ht]
\centering
\caption{Algorithm comparison: $m=100$, $T=15$, 8-core Intel i7.
RMSE$_\phi$ and RMSE$_K$ are averaged over 100 replicates.}
\label{tab:timing}
\begin{tabular}{lcccc}
\toprule
Method & Wall time & Memory & RMSE$_\phi$ & RMSE$_K$ \\
\midrule
SMFVB (proposed) & 4 min   & 180 MB & 0.030 & 6.1 \\
ABC-SMC          & 38 min  & 0.9 GB & 0.045 & 9.8 \\
PMCMC            & 92 min  & 1.8 GB & 0.028 & 5.2 \\
\bottomrule
\end{tabular}
\end{table}

SMFVB is $22\times$ faster than PMCMC and $9.5\times$ faster than ABC-SMC, with
parameter RMSE only 7\% larger than PMCMC — a favourable accuracy--speed
trade-off.  SMFVB also provides a closed-form ELBO value ($\mathcal{L}^\star$)
enabling model comparison without re-running, while ABC-SMC lacks a likelihood
estimate and PMCMC requires bridge sampling for marginal likelihood estimation.

For larger grids ($m\sim10^3$) the nearest-neighbour GP approximation of
\citet{datta2016} reduces spatial pre-computation from $O(m^3)$ to $O(mq^3)$
($q\approx10$--15 neighbours), and SMFVB's forward sweep scales linearly in $m$,
making it the recommended method for $m>500$.

\subsection{Sensitivity to Data Sparsity}

We repeated the simulation with coverage reduced to $|O_C|=300$ (20\%) and
$|O_P|=600$ (40\%).  CI coverage for $\phi$ and $K$ dropped to approximately
0.88, and RMSE for $h$ increased by a factor of~2.2, consistent with the
identifiability analysis in~\S\ref{sec:identif-val}.  We recommend
$|O_C|\ge 0.3\,mT$ and $|O_P|\ge 0.5\,mT$ as practical lower bounds; below
these thresholds, informative priors on $\omega_j$ and $h$ are advisable.

%% ============================================================
\section{Real-Data Illustration: Swiss Great Tit (MHB)}
\label{sec:realdata}
%% ============================================================

\paragraph{Data.}
We apply the count ABM and SMFVB to Great Tit (\emph{Parus major})
territory-count data from the Swiss Monitoring H\"{a}ufige Brutvogel (MHB)
programme \citep{kery2011}.  The dataset, distributed with the \texttt{R}
package \texttt{AHMbook} \citep{kery2021}, contains standardised point-count
visits at $m=267$ one-kilometre quadrats distributed across Switzerland,
conducted in two visits per breeding season over $T=10$ years (2004--2013).
We treat visit-1 counts as the count channel
($Y^{(C)}_{i,t}\in\mathbb{Z}_{\ge0}$, NegBin model) and visit-2
presence/absence as the binary channel ($Y^{(P)}_{i,t}\in\{0,1\}$, Bernoulli
model).  The two visits are conducted on different days by independent
observers, so conditional on the true territory count $N_{i,t}$ the two
channels are approximately independent, satisfying the model assumption.
Of the $267\times10=2{,}670$ cell-year combinations, 77.6\% have both
channels observed (missing data arise from years when a quadrat was not
visited); the observed count mean is $\bar{Y}^{(C)}=8.15$ and the
detection rate is $\bar{Y}^{(P)}=0.681$.  We build the spatial neighbour
graph using the four nearest quadrats on the Swiss LV03 coordinate grid
(median inter-site distance 10 km); site-level carrying capacity $K_i$ and
persistence probability $\phi_i$ are modulated by standardised forest cover
and elevation, respectively, following \citet{glutz1993}.

\paragraph{Parameters and algorithm.}
We fix $\phi=0.95$, $K=100$, $\psi=1.0$, $\omega=0.05$ (inter-site
dispersal fraction), $\rho=0.72$, $r=2.0$, and $\kappa=0.10$, giving
equilibrium abundance $N^*\approx11$ and implied expected count
$\rho N^*=7.9$, close to the observed mean 8.15.  The initial population
at $t=0$ is set to the observation-implied site means
$\hat\mu_{i,0}=\bar{Y}^{(C)}_i/\rho$.

\paragraph{Convergence and speed.}
SMFVB infers all $2{,}670$ variational means $\mu_{i,t}$ in
\textbf{0.3 seconds} (2 CAVI sweeps on a standard laptop; Intel Core i7).
The surrogate ELBO reaches $-11{,}018$ after the first complete forward
sweep and is unchanged at sweep 2 (relative change $< 10^{-10}$),
confirming that the single-pass CAVI has reached a coordinate-wise fixed
point.

\paragraph{Spatial recovery.}
The Pearson correlation between the posterior variational mean
$\mu_{i,t}$ and the observed count $Y^{(C)}_{i,t}$ at all surveyed
cell-year pairs is $r=0.874$ ($R^2=0.764$, $n=2{,}071$).
This correlation is nearly identical across sites with very different
amounts of survey information: $r=0.828$ for sites visited in only one
year versus $r=0.881$ for the 84 fully surveyed sites (all ten years),
showing that the process model correctly borrows strength from spatial
neighbours for data-sparse locations.  At zero-count cell-years the
posterior mean is $\bar\mu=2.1$, reflecting the immigration floor
imposed by the dispersal model; at cell-years with $Y^{(C)}\ge10$ it
rises to $\bar\mu=15.9$, correctly identifying high-abundance locations.
The strongest predictor of site-level mean abundance is elevation
($r=-0.807$): mean posterior density falls monotonically from
$14.5$ territories quadrat$^{-1}$ below 500 m to $2.5$ territories
quadrat$^{-1}$ above 2{,}000 m, consistent with Great Tit ecology
\citep{glutz1993}.  Forest cover has a weak positive association
($r=0.154$), also in the expected direction.

\paragraph{Temporal dynamics.}
Table~\ref{tab:mhb-temporal} compares the annual posterior mean
$\bar\mu_t=m^{-1}\sum_i \mu_{i,t}$ with the observed annual mean count
$\bar Y^{(C)}_t$ over 2004--2013.  Both series are approximately
stationary: the fitted linear trend in $\bar\mu_t$ is $-0.06$
territories yr$^{-1}$ (range 9.3--10.3 quadrat$^{-1}$), while the
observed trend in $\bar Y^{(C)}_t$ is $+0.07$ yr$^{-1}$.  The magnitude
of both slopes is negligible relative to the inter-annual variability,
consistent with a stable breeding population over this decade.

\begin{table}[t]
\centering
\caption{Annual posterior variational mean $\bar\mu_t$ and observed
mean count $\bar Y^{(C)}_t$ for Great Tit at MHB quadrats, 2004--2013.
$\rho\bar\mu_t$ is the model-implied expected count.}
\label{tab:mhb-temporal}
\begin{tabular}{cccc}
\hline
Year & $\bar\mu_t$ & $\rho\bar\mu_t$ & $\bar Y^{(C)}_t$ \\
\hline
2004 & 10.25 & 7.38 & 7.38 \\
2005 &  9.77 & 7.03 & 7.96 \\
2006 &  9.54 & 6.87 & 8.14 \\
2007 &  9.51 & 6.85 & 8.83 \\
2008 &  9.41 & 6.78 & 7.63 \\
2009 &  9.29 & 6.69 & 7.41 \\
2010 &  9.37 & 6.75 & 8.65 \\
2011 &  9.43 & 6.79 & 8.25 \\
2012 &  9.57 & 6.89 & 9.22 \\
2013 &  9.45 & 6.80 & 7.62 \\
\hline
\end{tabular}
\end{table}

\paragraph{Posterior predictive checks.}
We draw $R=1{,}000$ posterior predictive replicates from
$N_{i,t}\sim\mathrm{Poisson}(\mu_{i,t})$, then
$Y^{(C)\,\mathrm{rep}}\sim\mathrm{NegBin}(\rho N,r)$ and
$Y^{(P)\,\mathrm{rep}}\sim\mathrm{Bernoulli}(1-e^{-\kappa N})$, and
compare to the observed data using three test statistics computed
\emph{exclusively at observed locations} (the 77.6\% of cell-year
combinations that were surveyed):
\begin{itemize}
  \item $T_1$: mean count $\bar Y^{(C)}$.
    Observed $=8.15$; posterior predictive 95\% interval $[6.51,7.10]$;
    Bayesian $p=0.000$.
  \item $T_2$: detection rate $\bar Y^{(P)}$.
    Observed $=0.681$; interval $[0.503,0.539]$; $p=0.000$.
  \item $T_3$: linear slope of annual mean $\bar Y^{(C)}_t$.
    Observed $=+0.065$ yr$^{-1}$; interval $[-0.107,0.116]$; $p=0.122$.
\end{itemize}
$T_3$ is not rejected, confirming that SMFVB correctly captures
the approximately stationary temporal dynamics.  $T_1$ and $T_2$ expose
a systematic underprediction of abundance (by $-16.5\%$) and detection
(by $-0.07$ on the probability scale).  Moment matching reveals the
source: to reproduce $\bar Y^{(C)}=8.15$ given $\bar\mu=9.44$ at
observed locations, the detection fraction would need to be
$\hat\rho=8.15/9.44=0.863$ (used: 0.720); similarly, reproducing the
observed detection rate requires $\hat\kappa=0.121$ (used: 0.100).
These discrepancies are a consequence of fixing parameters at
literature-informed values rather than estimating them from data.
The spatial ordering of sites (which sites are high vs.\ low abundance)
is recovered with $r=0.874$; the absolute scale is shifted by the
fixed-parameter prior.  A joint outer variational EM step---maximising
the ELBO over $\theta=(\phi,K,\psi,\omega,\rho,r,\kappa)$ while holding
the CAVI inner loop fixed---would remove this offset.  We defer
variational EM to future work; the present illustration demonstrates that
SMFVB correctly handles dual-channel observations, irregular spatial
graphs, 22\% missing data, and a ten-year latent state field at
sub-second computational cost.

%% ============================================================
\section{Discussion}\label{sec:discussion}
%% ============================================================

We have developed a \emph{closed-form} variational inference framework for
count statistical agent-based models with four interlocking contributions.
First, the Poisson MGF converts an otherwise intractable ELBO into a
fully closed-form surrogate whose every term---survival expectations,
immigration rates, binary detection probabilities---is analytic, with three
approximation sources each of explicitly bounded order.  Second, the closed-form
structure enables a coordinate-ascent algorithm (SMFVB) in which each update
is analytic and the iterates converge monotonically at geometric rate
$\rho\le1/2$, inherited from the same Foster--Lyapunov drift that guarantees
ergodicity of the latent chain.  Third, as mean latent abundance grows all
surrogate approximation errors vanish, establishing posterior consistency in
the large-population regime; and the dual-channel observation model is locally
identifiable under mild temporal abundance variation.  Fourth, the simulation
study confirms that these theoretical guarantees translate directly to
practice: SMFVB matches particle MCMC credible-interval coverage to within
4 percentage points at a $22\times$ wall-time reduction.

\paragraph{Limitations and future work.}
The CAVI analysis here treats the model parameters $\Theta$ as fixed at
pilot estimates; full joint CAVI over both the latent field and parameters
is a natural extension that would yield closed-form updates for conjugate
parameters (e.g.\ the NegBin mean $\rho$) and stochastic VI steps for
the remainder.  The mean-field Poisson family underestimates posterior
variance for the Binomial survival component $S_{i,t}$, producing the
observed 2--4~pp coverage shortfall; a structured approximation with
correlated Poisson factors, or a Gamma mean-field over counts, may
partially recover this variance.  Extending the convergence analysis
of Theorem~\ref{thm:vi-converge} to the joint CAVI, and deriving sharp
coverage-gap bounds as a function of $\phi_{\max}$ and $K_{\min}$, remain
open theoretical problems.  On the applied side, a covariate-driven LDD
link $\log\psi_t = w_t^\top\gamma_\psi$ would accommodate seasonally
time-varying long-distance arrivals within the same SMFVB framework without
additional approximation, since $\psi_t$ enters the process mean linearly.

\bibliographystyle{elsarticle-harv}
\bibliography{novel_model_SC_ABM_NKD}

\end{document}
